---
title: "Main Project Document"
author: "Alan Corp & Charlie Argust"
date: "2025-11-21"
output: html_document
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(fpp3)
library(Metrics) # rmse calculation
library(glue) # formatted strings
library(fable)
library(feasts)
library(dplyr)
library(Metrics)
```

## Australian Pharmaceutical Benefits Scheme (PBS) - Time Series Analysis

### Project Description and Data

The PBS dataset from the `fpp3` package contains information regarding Australian's 
government subsidy scheme for prescription drugs, containing monthly observations 
from July 1991 to June 2008. This scheme enables Australian citizens to obtain 
prescription medications at reduced cost, with the Australian government 
subsidizing the difference. This project aims to explore the time series 
patterns of prescription costs and exogenous variables that correlate with these 
observed patterns. Through an analysis of this data, we can build time series 
models to forecast future subsidy demand and in theory help the government 
financially plan for subsequent years.

The data is initially aggregated by Concession, Type, ATC1, and ATC2, which are described below:

- `Concession`: patient type
    * Concessional - Patients including pensioners, unemployed, dependents, and 
    other card holders that also have a Medicare card. Concessional patients are 
    subject to entitlements, including lower co-payments.
    * General - Patients that have a Medicare card, but do not meet the above
    criteria.
- `Type`: payment type
    * Co-payments - Fixed amount that a customer pays out of pocket. Co-payments are
    made until a patient's total script expenditure hits a threshold, at which point
    safety net subsidies provide additional discounts.
    * Safety net - Prescriptions obtained after a patient hits a script expenditure 
    threshold
    
The ATC (Anatomical Therapeutic Chemical) system is a hierarchical method of classifying 
drugs based on their properties, each level increasing in specificity [1]. The `PBS` 
dataset provides the first 2 levels:

- `ATC1`: Anatomical main group
    * e.g. 'Cardiovascular system', 'Nervous system', 'Alimentary tract & metabolism'
- `ATC2`: Therapeutic subgroup
    * e.g. 'Drugs for acid-related disorders', 'Psychoanaleptics', 'Antidiabetic therapy'

The scope of this project includes all antidiabetic therapy drug prescriptions (ATC2 == 'A10'). Hence, we can disaggregate the data to include all `Concession` types and payment (`Type`) types.

### Data Exploration

```{r}
PBS%>%
  filter(year(Month) == 2008)%>%
  index_by(year = year(Month))%>%
  group_by(ATC1, ATC1_desc)%>%
  summarise(TotalScripts = sum(Scripts),
            TotalCost = sum(Cost))%>%
  arrange(desc(TotalScripts))%>%
  head(5)
```

Prescriptions classified as aiding the Cardiovascular system, nervous system, and Alimentary tract and metabolism made up the largest share of scripts in 2008. 

```{r}
PBS%>%
  as_tibble()%>%
  filter(ATC1 %in% c('C', 'N', 'A'))%>%
  group_by(ATC2_desc)%>%
  summarise(Scripts = sum(Scripts))%>%
  arrange(desc(Scripts))
```

### Antidiabetic Therapy

```{r, filtering and aggregating antidiabetic for antidiabetic scripts}
antidiabetic_med <- PBS%>%
  filter(ATC2 == 'A10')%>%
  group_by(ATC2, ATC2_desc)%>%
  summarise(TotalPrescriptions = sum(Scripts),
            TotalSubsidy = sum(Cost))

antidiabetic_med
```

```{r}
summary(antidiabetic_med[,c('TotalPrescriptions', 'TotalSubsidy')])
```

```{r reducing scale of values}
antidiabetic_med <- antidiabetic_med%>%
  mutate(TotalPrescriptions = TotalPrescriptions / 1e+3, # transform scripts to thousands
         TotalSubsidy = TotalSubsidy / 1e+6) # subsidy to millions
  
mean(antidiabetic_med$TotalPrescriptions); mean(antidiabetic_med$TotalSubsidy)
```

```{r}
cor(antidiabetic_med$TotalPrescriptions, antidiabetic_med$TotalSubsidy)
```

As one would expect, the number of dispensed prescriptions is highly correlated
with the subsidy. This project will be mainly focusing on the subsidy.

### Visualizing Time Series

```{r}
antidiabetic_med%>%
  autoplot(TotalPrescriptions) # log transformation works to stabilize variance
```

```{r}
theme_set(theme_get() + theme(text = element_text(family = 'sans')))

antidiabetic_med%>%
  ggplot(aes(Month, TotalSubsidy)) + geom_line(size = 0.7) +
  labs(title = 'PBS monthly subsidy', 
       subtitle = 'Antidiabetic therapy prescriptions',
       y = '$AUD (Millions)') +
  theme(text = element_text(size = 12))
```
The time series plot of total subsidies provided for antidiabetic therapy drug prescription reveals strong trend and seasonality from 1991 to 2008. Monthly subsidies show a steady increase, with a slight quadratic trend indicating a compounding effect. Subsidies also show a clear peak in January, which is followed by an immediate descent into a trough in February. The seasonal plot below confirms this:

```{r}
antidiabetic_med%>%
  select(Month, TotalSubsidy)%>%
  gg_subseries()
```

Considering the need for antidiabetic drugs doesn't vary from month-to-month, one might wonder why seasonality is present in this data. The data represents the frequency with which prescriptions are dispensed, not the frequency that the drugs are actually used. According to the research I've found, peaks in prescriptions dispenses and total subsidies typically appear in December rather than January [2]. This is due to the safety net being reached, so December is the month where patients can stockpile their prescriptions at a cheaper price (hence a higher government subsidy) before it resets at the beginning of a new year. I suspect that the 1-month discrepancy between researched PBS seasonality and the seasonality present in the data shown previously is due to a lag from the date that pharmacies dispense the drug and the date that a claim for reimbursement is processed.

### 2008 Shift in Insulin Usage

```{r}
monthly_scripts <- antidiabetic_med%>%
  index_by(year = year(Month))%>%
  summarise(scripts_per_month = sum(TotalPrescriptions) / length(Month))

monthly_scripts
```

The average number of prescriptions subsidized by the PBS per month increases continuously until 2008. This aligns with the general increasing trend in diabetes prevalence, however diabetes prevalence does not decrease in 2008. Additionally, the mortality rate in Australia for diabetes patients reached a 20-year high. 

```{r}
diabetes_data <- as.data.frame(
  list(
    years = seq(2000, 2013, by = 1),
    prevalence_pct = c(2.4, 2.7, 2.9, 3.1, 3.3, 3.5, 3.6, 3.8, 4.0, 4.1, 4.2, 4.3, 4.4, 4.4),
    deaths_per_100k = c(54.3, 53.5, 57.8, 56.2, 56.5, 55.4, 58.3, 57.8, 61.6,59.3, 57.6, 58.8, 57.0, 55.7)
    )
)

diabetes_long <- pivot_longer(diabetes_data,
             cols = c(prevalence_pct, deaths_per_100k),
             names_to = 'statistic',
             values_to = 'value')

ggplot(diabetes_long, aes(x = years, y = value)) + 
  geom_line(color = 'blue') + 
  geom_point(color = 'red') + 
  facet_wrap(~statistic, ncol = 1, scales = 'free_y')
```

```{r, warning=FALSE}
monthly_scripts$prevalence <- c(NA,NA,NA,NA,NA,NA,NA,NA,NA, 2.4, 2.7, 2.9, 3.1, 3.3, 3.5, 3.6, 3.8, 4.0)

data_long <- pivot_longer(monthly_scripts,
             cols = c(scripts_per_month, prevalence),
             names_to = 'statistic',
             values_to = 'value')

data_long$statistic <- recode(
  data_long$statistic,
  scripts_per_month = "Scripts per Month",
  prevalence = "Diabetes Prevalence (%)"
)


ggplot(data_long, aes(x = year, y = value)) + 
  geom_line(color = 'blue') + 
  geom_point(color = 'red') + 
  facet_wrap(~statistic, ncol = 2, scales = 'free_y') + 
  theme(strip.text = element_text(
    face = "bold",     # bold facet titles
    size = 12          # increase font size
  ))

```
# https://www.aihw.gov.au/reports/diabetes/diabetes/contents/explore-the-data

```{r, eval = FALSE}
ndss_data <- as.data.frame(
  list(
    fiscal_year <- seq(1992, 2008, by = 1),
    non_insulin_using = c()
  )
)

#non_insulin_using <- c(29476,  42643,  58,300 76,772  98,810 128,162  161,678  202,549 246,805  298,599  353,825  409,150 466,409 525,365 588,580 622,340 577,567 
```


### Autocorrelation plot

```{r}
autoplot(ACF(antidiabetic_med, TotalSubsidy)) +
  labs(title = 'Autocorrelation of monthly subsidy', 
       subtitle = 'Antidiabetic therapy prescriptions',
       x = 'Lag [1 Month]',
       y = 'ACF') +
  theme(text = element_text(size = 12))
```


### Fitting Models

Before fitting models, we will split the data into 90% train and 10% test, so we can compare the model forecasts against the true values.

```{r}
subsidy <- antidiabetic_med%>%
  ungroup(ATC2)%>%
  index_by(Month)%>%
  select(Month, TotalSubsidy)

subsidy
```


```{r splitting into train and test}
split_idx <- as.integer(0.9 * nrow(subsidy))

train = subsidy[1:(split_idx - 1),]
test = subsidy[split_idx:nrow(subsidy),]

nrow(train); nrow(test)
```

#### Simple Baseline Models

```{r, message=FALSE}
baseline_fits <- train%>%
  model(
    naive = NAIVE(TotalSubsidy),
    snaive = SNAIVE(TotalSubsidy),
    mean = MEAN(TotalSubsidy),
    drift = NAIVE(TotalSubsidy ~ drift())
  )

baseline_forecasts <- baseline_fits%>%
  forecast(h = 22)
```

```{r}
baseline_forecasts%>%
  ggplot(aes(Month, .mean, color = .model)) + geom_line(size = 0.8) +
  autolayer(train, TotalSubsidy, size = 0.8) +
  autolayer(test, TotalSubsidy, alpha = 0.5, linetype = 'dashed', size = 0.6, color = 'black') + 
  labs(title = 'Simple forecasts of monthly subsidy', 
       subtitle = 'Antidiabetic therapy prescriptions',
       y = '$AUD (Millions)') + guides(colour = guide_legend(title = "Forecast")) +
  theme(text = element_text(size = 12))
```

```{r}
models <- baseline_forecasts[,c('.model', '.mean')]
actual_values = test$TotalSubsidy

for (model in unique(models$.model)) {
  
  current_model <- models%>%
    filter(.model == model)
  
  error = rmse(actual_values, current_model$.mean)
    print(glue("{model} forecast | RMSE: {error}"))
}
```

#### Regression Models

```{r}
regression_fit <- train%>%
  model(
    TSLM(log(TotalSubsidy) ~ trend() + season())
  )

regression_fit%>%
  forecast(h = 22)%>%
  autoplot(train, level = NULL) +
  autolayer(test, color = 'red', size = 0.75, linetype = 'dashed')
```

#### Check Stationarity

```{r}
# Stationarity tests — KPSS and ADF
train %>% features(TotalSubsidy, unitroot_kpss)
train %>% features(TotalSubsidy, unitroot_ndiffs)
train %>% features(TotalSubsidy, unitroot_nsdiffs)

```

The KPSS test suggests non-stationarity (p<.05). The ndiffs and the nsdiffs indicate a first-order differencing is needed but seasonal differencing is not required. 

#### Seasonal and Non-Seasonal Differencing

```{r, warning=FALSE}
# reconstruct as a monthly tsibble
subsidy_ts <- PBS %>%
  filter(ATC2 == "A10") %>%
  index_by(Month) %>%
  summarise(TotalSubsidy = sum(Cost) / 1e6) %>%   # millions scale
  as_tsibble(index = Month)

subsidy_ts %>%
  model(STL(TotalSubsidy)) %>%
  components() %>%
  autoplot()
```

#### ACF/PACF Plots

```{r}
diff_subsidy <- subsidy_ts %>%
  mutate(
    diff_1 = difference(TotalSubsidy),
    diff_12 = difference(TotalSubsidy, lag = 12),
    diff_both = difference(diff_1, lag = 12)
  ) %>%
  filter(!is.na(diff_both))


# ACF/PACF for raw series
ACF(train, TotalSubsidy) %>% autoplot()
PACF(train, TotalSubsidy) %>% autoplot()

# ACF/PACF after differencing
ACF(diff_subsidy, diff_both) %>% autoplot()
PACF(diff_subsidy, diff_both) %>% autoplot()
```
The ACF shows strong seasonal spikes at lag 12 and slow decay, supporting SARIMA.
PACF suggests possible AR(1) or AR(2) structure after differencing.

#### Fit ARIMA/SARIMA Models

```{r}
# Fit an automatically selected ARIMA
fit_auto <- train %>%
  model(ARIMA(TotalSubsidy))

report(fit_auto)

# Fit a manually specified SARIMA based on ACF/PACF
fit_manual <- train %>%
  model(
    sarima = ARIMA(TotalSubsidy ~ pdq(1,1,1) + PDQ(0,1,1))
  )

report(fit_manual)

```

#### Residual Diagnostics

```{r}
fit_auto <- subsidy_ts %>%
  model(ARIMA(TotalSubsidy))
fit_manual <- subsidy_ts %>%
  model(ARIMA(TotalSubsidy ~ pdq(0,1,1) + PDQ(0,1,1)))


# Residual diagnostics for both models
gg_tsresiduals(fit_auto)
gg_tsresiduals(fit_manual)

# Ljung-Box tests

# Automatic ARIMA
augment(fit_auto) %>%
  features(.resid, ljung_box, lag = 24, dof = 2)

# Manual ARIMA
augment(fit_manual) %>%
  features(.resid, ljung_box, lag = 24, dof = 2)

```

#### Forecasting for ARIMA Models

```{r}
fc_auto <- fit_auto %>% forecast(h = nrow(test))
fc_manual <- fit_manual %>% forecast(h = nrow(test))

autoplot(fc_auto, train) +
  autolayer(test, color="red", linetype="dashed")

```
```{r}
# 1. Refit ARIMA models using the TRAIN data ONLY
fit_auto <- train %>%
  model(ARIMA(TotalSubsidy))

fit_manual <- train %>%
  model(ARIMA(TotalSubsidy ~ pdq(1,1,1) + PDQ(0,1,1)))

# 2. Forecast horizon = size of test set
h <- nrow(test)

fc_auto   <- fit_auto %>% forecast(h = h)
fc_manual <- fit_manual %>% forecast(h = h)

# 3. Combine all forecasts into fc_all
fc_all <- bind_rows(
  baseline_forecasts %>% mutate(.model = as.character(.model)),
  fc_auto        %>% mutate(.model = "auto_arima"),
  fc_manual      %>% mutate(.model = "manual_arima")
)

```

#### Compare Model Accuracies

```{r}
actual_values <- test$TotalSubsidy

#list of unique models in forecast object
model_names <- unique(fc_all$.model)

# Create an empty data frame to store accuracy results
accuracy_results <- data.frame(
  Model = character(),
  RMSE = numeric(),
  MAE = numeric(),
  MAPE = numeric(),
  stringsAsFactors = FALSE
)

# Loop over each model to calculate metrics
for (model_name in model_names) {
  predicted_values <- fc_all %>%
    filter(.model == model_name) %>%
    pull(.mean)
  
  # Accuracy metrics
  rmse_val <- rmse(actual_values, predicted_values)
  mae_val  <- mae(actual_values, predicted_values)
  mape_val <- mape(actual_values, predicted_values)
  
  # Add results to the data frame
  accuracy_results <- rbind(accuracy_results,
                            data.frame(Model = model_name,
                                       RMSE = rmse_val,
                                       MAE = mae_val,
                                       MAPE = mape_val))
}

accuracy_results
 

```

The best model is drift with the lowest RMSE/MAPE for this monthly seasonal data.

#### Plotting Final Model/Residuals

```{r}
#final_fit <- fit_manual  # choose whichever has best accuracy

#final_fit %>% forecast(h = 24) %>% autoplot(subsidy)
#gg_tsresiduals(final_fit %>% select(`ARIMA(TotalSubsidy ~ pdq(1, 1, 1) + PDQ(0, 1, 1))`))





```

#### Additional Seasonal Decomposition

```{r}
#subsidy %>%
#  model(STL(TotalSubsidy ~ season(window = "periodic"))) %>%
#  components() %>%
#  autoplot()
```


## References

[1] ATC classification system: https://atcddd.fhi.no/atc/structure_and_principles/

- About PBS: https://www.pbs.gov.au/info/about-the-pbs

- Medicare: https://www.health.gov.au/topics/medicare/about

- page 31: https://www.pbs.gov.au/statistics/2012-2013-files/expenditure-and-prescriptions-12-months-to-30-06-2013.pdf

- seasonality in PBS data and data description: 
[2] https://pmc.ncbi.nlm.nih.gov/articles/PMC4630883/